\documentclass[12pt,a4paper]{report}
\usepackage{graphicx}
\usepackage[space]{grffile}
\usepackage{latexsym}
\usepackage{textcomp}
\usepackage{longtable}
\usepackage{multirow,booktabs}
\usepackage{amsfonts,amsmath,amssymb}
\usepackage[round]{natbib}
\usepackage{url}
\usepackage[backref]{hyperref}
\hypersetup{colorlinks=true,citecolor=blue,pdfborder={0 0 0}}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{lscape}
\usepackage[margin=1in]{geometry}
\usepackage{parskip}
\setcitestyle{aysep={}}
\usepackage[autostyle]{csquotes}
\usepackage[none]{hyphenat}
\usepackage{graphicx,wrapfig,lipsum}
\usepackage{textgreek}
\usepackage{varioref}
\usepackage[tikz]{bclogo}
\usepackage{tabu}
\usepackage{pdflscape}
\usepackage{rotating}
\usepackage{microtype}


\usepackage{imakeidx}
\makeindex%[columns=3, title=Alphabetical Index, intoc, options= -s example_style.ist]


\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage{verbatim}


\input{Alias}


%\usepackage{setspace}
%\doublespacing
% or:
%\onehalfspacing %STEVE'S PREFERENCE
%\linespread{1.1}

\title{{%
		Individual and Community Development: \\
		\Large The Vision, Perceptions and Role of Early Adolescent Youth}}

\author{Anneleis Humphries \\
	Melbourne Graduate School of Education \\
	University of Melbourne \\
	Supervisors: Assoc. Prof Helen Cahill \& Dr Kylie Smith}

\date{\today}
\bibliographystyle{anneleis-agsm}
\begin{document}
	
%	

	%
	
	\maketitle 
	\hypersetup{linkcolor=blue} 
	\tableofcontents
	\newpage
	%\input{chapters/Ch01intro}
	\chapter{Intro}
	\index{Research Questions}
This research seeks to answer the following questions:
		\begin{itemize} 
			\item How do young people (aged 12-15) understand their own capacity to contribute to their communities?
			\begin{itemize}
				\item How do young people, in particular those in early adolescence, view community development? 
				\item What do they understand to be the role of young people in the process of community development?
				\item What do young people see as important in community building processes? What is their vision? How do they envision achieving this?
				\item What do they believe would motivate young people to get involved in community development activities?
			\end{itemize}
		\end{itemize}
	
	\input{chapters/Ch02PeriodofYouth}
	\input{chapters/Ch03YouthNCommunityBuilding}
	\input{chapters/Ch04LearningandCB}
	%\input{chapters/Ch05methods}
	%\input{chapters/Ch07conclusion} 
	%\input{chapters/Ch00Overview}
	\printindex
	\bibliography{mendeley.bib}
	\addcontentsline{toc}{chapter}{Bibliography}
%	\bibliography{../bibliography/PHDReferences}
\end{document}
